\documentclass[12pt]{article}
\usepackage{pmmeta}
\pmcanonicalname{ComplementarySubspace}
\pmcreated{2013-03-22 12:52:16}
\pmmodified{2013-03-22 12:52:16}
\pmowner{rmilson}{146}
\pmmodifier{rmilson}{146}
\pmtitle{complementary subspace}
\pmrecord{11}{33209}
\pmprivacy{1}
\pmauthor{rmilson}{146}
\pmtype{Definition}
\pmcomment{trigger rebuild}
\pmclassification{msc}{15A03}
\pmsynonym{algebraic complement}{ComplementarySubspace}
\pmdefines{complementary}
\pmdefines{direct sum}
\pmdefines{decomposition}
\pmdefines{orthogonal complement}

\endmetadata

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\newcommand{\reals}{\mathbb{R}}
\newcommand{\natnums}{\mathbb{N}}
\newcommand{\cnums}{\mathbb{C}}
\newcommand{\znums}{\mathbb{Z}}
\newcommand{\lp}{\left(}
\newcommand{\rp}{\right)}
\newcommand{\lb}{\left[}
\newcommand{\rb}{\right]}
\newcommand{\supth}{^{\text{th}}}
\newtheorem{proposition}{Proposition}
\newtheorem{definition}[proposition]{Definition}
\newcommand{\nl}[1]{{\PMlinkescapetext{{#1}}}}
\newcommand{\pln}[2]{{\PMlinkname{{#1}}{#2}}}
\newcommand{\kf}{\mathbb{K}}
\begin{document}
\paragraph{Direct sum decomposition.}
Let $U$ be a vector space, and $V,W\subset U$ subspaces.  We say that
$V$ and $W$ span $U$, and write
$$U=V+W$$ if 
every $u\in U$ can be expressed as a sum
$$u=v+w$$
for some $v\in V$ and $w\in W$.

If in addition, such a decomposition is unique for all $u\in U$, or
equivalently if 
$$V\cap W=\{ 0\},$$
then we say that $V$ and $W$ form a \emph{direct sum decomposition} of  $U$
and write
$$U=V\oplus W.$$
In such circumstances, we also say that $V$ and $W$
are \emph{complementary} subspaces, and also say that $W$ is an \emph{algebraic complement} of $V$.

Here is useful  characterization of complementary subspaces if $U$ is
finite-dimensional. 
\begin{proposition}
  Let $U, V, W$ be as above, and suppose that $U$ is
  finite-dimensional.  The subspaces $V$ and $W$ are complementary if
  and only if for every basis  $v_1,\ldots, v_m$  of $V$ and
  every basis
  $w_1,\ldots,w_n$  of $W$, the combined list
  $$v_1,\ldots,v_m,w_1,\ldots,w_n$$
  is a basis of $U$.
\end{proposition}

\textbf{Remarks}.
\begin{itemize}
\item
Since every linearly independent subset of a vector space can be extended to a basis, every subspace has a complement, and the complement is necessarily unique.
\item
Also, direct sum decompositions of a vector space $U$ are in a one-to correspondence fashion with projections on $U$.
\end{itemize}

\paragraph{Orthogonal decomposition.}
Specializing somewhat, suppose that the ground field $\kf$ is either
the real or complex numbers, and that $U$ is either an inner product
space or a unitary space, i.e. $U$ comes equipped with a
positive-definite inner product
$$\langle,\rangle:U\times U\rightarrow \kf.$$
In such circumstances,
for every subspace $V\subset U$ we define the orthogonal complement of
$V$, denoted by $V^\perp$ to be the subspace
$$V^\perp = \{ u\in U: \langle v,u\rangle = 0,\text{ for all }
v\in V\}.$$
\begin{proposition}
  Suppose that $U$ is finite-dimensional and $V\subset U$ a subspace.
  Then,  $V$ and its orthogonal
  complement $V^\perp$ determine a direct sum decomposition  of $U$.
  \end{proposition}
  
  Note: the Proposition is false if either the finite-dimensionality
  or the positive-definiteness assumptions are violated.
%%%%%
%%%%%
\end{document}
